\Lecture{Introduction}

Important links: 
\begin{itemize}
    \item Link to course book pdf: \href{https://dpvipracollege.ac.in/wp-content/uploads/2023/01/John-E.-Hopcroft-Rajeev-Motwani-Jeffrey-D.-Ullman-Introduction-to-Automata-Theory-Languages-and-Computations-Prentice-Hall-2006.pdf}{here}
    \item Solution manual for exercises: \href{http://infolab.stanford.edu/~ullman/ialc.html#solutions}{here}
\end{itemize}

\subsection{Proofs}

\subsubsection{Deductive proofs} 
A deductive proof is a logical sequence of statements from some initial statement (\textit{hypothesis} $H$) to a conclusion statement ($C$). The general form of a deductive theorem is usually 
\begin{itemize}
    \item "if $H$ then $C$"
\end{itemize}

If we do no know where to begin with a proof it is useful to reduce all the terms in the hypothesis to their definitions. 
\begin{example} 
    Definitions: 
    \begin{itemize}
        \item A set $S$ is finite if there exists some integer such $u$ that $|S| = u$.
    \end{itemize}
    
    \textbf{Theorem:} Let $S,T$ be some finite sets. If $S \cup T = U$ and $S \cap T = \emptyset$ then $|U| = |S| + |T|$.

    Proof: Trivial from definitions.
\end{example}

Two sided proofs come in the form "A if and only if B". These have to be proven by two proofs 
\begin{itemize}
    \item The "If A then B" part 
    \item The "If B then A" part
\end{itemize}

It is sometimes useful to prove the \textit{contrapositive} statement. A contrapositive statement exists for every "If H then C" statement and comes in the form "If not C then not H". Both forms are logically equivalent.

\subsubsection{Inductive proofs}
Inductive proofs often deal with sequences, for example we might want to prove that some $S(n)$ holds for all $n$ greater than some $k$.

Induction is not limited to integers however. Structural induction deals with proving inductive statements over any recursively defined structure. That is, if we have some recursive definition for some structure $X$ then, we can prove statements in the form $S(X)$ with the same inductive principles as for integers.

Mutual induction is when we cannot prove the statement $S(n)$ alone and, instead have to prove a group of statements $S_1(n), S_2(n), \dots,S_k(n)$ together. Here each of the statements must be proved separately in the basis and in the inductive step. If the statements are of the two sided form then both sides have to be proved for each statement.

\subsection{Central definitions}
The first lecture introduces some central definitions relevant to automata theory and formal languages. \\

\begin{definition}[Alphabet] \label{def:alphabet}
    An alphabet is a finite, nonempty set of symbols, which by convention it is denoted by $\Sigma$.
\end{definition}
\begin{example}
    An alphabet can be almost any finite nonempty set. Some common ones include: 
    \begin{itemize}
        \item The binary alphabet $\Sigma = \set{0,1}$
        \item The set of all lowercase letters $\Sigma = \set{a,b,\dots, z}$
        \item The set of ASCII characters
    \end{itemize}
\end{example}

\begin{definition}[String] \label{def:string}
    A string (or word) is a finite sequence of symbols from some alphabet $\Sigma$
\end{definition}

\begin{example}
  $1010$ is a string from the binary alphabet $\Sigma = \set{0,1}$.
\end{example}

\begin{definition}[The empty string] \label{def:empty-string}
    The empty string is a string with zero occurrences of symbols. It can be chosen from any alphabet and, is by convention denoted as $\epsilon$.
\end{definition}

\begin{definition}[The set of all strings]
    The set of all strings over some alphabet $\Sigma$ is denoted by $\Sigma^*$. It is simply the powerset over $\Sigma$.
\end{definition}

In the following definitions and examples consider strings defined over the alphabet $\Sigma = \set{a}$, belonging to the language (see \ref{def:language}) defined as $$\sigma ::= \epsilon | a\sigma$$

\begin{definition}[The length of a string] \label{def:string-length}
    The length of a string can be inductively defined as follows 
    \begin{align*}
        |\epsilon| &\defeq 0 \\
        |ax| &\defeq 1 + |x|
    \end{align*}
    ($|x|$ is the conventional notation for the length of a string)
\end{definition}

\begin{definition}[String concatenation] \label{def:string-concatenation}
    Let $x,y$ be strings, then $xy$ or $x \cdot y$ denotes the concatenation of $x$ and $y$. It is inductively defined as 
    \begin{align*}
        \epsilon \cdot x &\defeq x \\
        (a \cdot x) \cdot y &\defeq a \cdot (x \cdot y)
    \end{align*}
\end{definition}

Note that for any string $w$ the equation $\epsilon w = w\epsilon = w$ holds, this means that $\epsilon$ is the identity element for concatenation. 

\begin{definition}[Language] \label{def:language}
    If $\Sigma$ is an alphabet and $L\subseteq \Sigma^*$ then $L$ is a language over $\Sigma$. 
\end{definition}

\begin{example}
    Let $\Sigma = \set{0,1}$ then $$L = \set{\epsilon, 1, 0, 11, 00, 01, 10}$$ is a language over $\Sigma$. For every alphabet the smallest possible language is $L = \emptyset$ and the largest possible $L = \Sigma^*$.
\end{example}

\begin{definition}[Language concatenation] \label{def:language-concatenation}
Let $L_1, L_2 \subseteq \Sigma^*$ for some alphabet $\Sigma$ then their concatenation is defined as $$L_1 \cdot L_2 \defeq \setdef{x_1 \cdot x_2}{ x_1 \in L_1, x_2\in L_2}$$ 
\end{definition}

A language can be concatenated with itself, resembling powers of a language. Let $L \subseteq \Sigma^*$ then $L^k$ can be inductively defined as 
\begin{align*}
    L^0 &\defeq \set{\epsilon} \\
    L^{i+1} &\defeq L \cdot L^i
\end{align*}
Notice that $L^0 \defeq \set{\epsilon}$, an alternative would be to define $L^0 \defeq \emptyset$ however, in that case $L^k$ would reduce to $\emptyset$ and $L^1 \neq L$ since $L\cdot \emptyset = \emptyset$. Hence $\set{\epsilon}$ is the \textit{identity} for language concatenation and, $\emptyset$ is an \textit{absorbing} element.

\begin{definition}[Kleene star] \label{def:kleene-star}
    Let $L\subseteq \Sigma^*$ for some alphabet $\Sigma$, then $L^*$ and $L^+$ are defined as 
    \begin{align*}
        L^* &\defeq \bigcup_{i \geq0} L^i \\
        L^+ &\defeq \bigcup_{i > 0} L ^i
    \end{align*}
\end{definition}

\subsubsection{Problems}
In automata theory problems are really the same thing as languages. A problem is the question of deciding whether a given string is a member of some language. Hence, in automata theory we are generally only concerned with decision problems. Whether a decision problem can be solved efficiently is always a lower bound to the corresponding construction problem.